Classical Simulation of Quantum Error Correction in a Fibonacci Anyon Code

### 2015-06-10 fib2.tex
many more edits.

%------------------------------------------------------------------------------------------------------------%
%\paragraph{Modeling Fibonacci anyons.}
%\paragraph{Topological quantum field theory.}
\paragraph{Topological model.}

@@

We consider a 
2-dimensional system with
point-like excitations or quasiparticles known as anyons.
These anyons show exotic statistics under exchange.
The worldlines of these exchanges form \emph{braids}
in (2+1)-dimensions
which act unitarily on (the Hilbert space of) degrees of freedom known as the \emph{fusion space.}
To access this space we define
the topological observables of the system: these
are total charge measurements within regions bounded by closed loops (Wilson loops).
%The algebraic properties of these processes
These dynamics and observables obey algebraic rules
given by a unitary modular tensor category~\cite{Wang2010b}. % (UMTC).

@@

From the perspective of
\emph{topological quantum field theory}
anyons are modelled
as punctures in a surface~\cite{Pfeifer2014}, and it is not
the anyons themselves that braid, but the manifold in which the anyons reside 
becomes deformed (twisted) by these braid moves.
This is the perspective we adopt.
In particular, this shows the action of braiding
on arbitrary observables (Wilson loops) as
these deform with the manifold.

The state of a collection of anyons
can be described by the observables of the system.
The difference between abelian and non-abelian
anyon theories is that the outcomes of measurements in an abelian theory
are determined uniquely particle content only.
That is, the fusion space is one-dimensional for abelian anyons, and generally larger for non-abelian anyons.
%An example is Fibonacci: $\tau\times\tau=\tau+\vac$
%\emph{picture here?}
%specify a state requires nesting multiple wilson lines.
In particular, we consider a system supporting (non-abelian) Fibonacci
anyon excitations, denoted by $\tau.$
Two such anyons can have total charge
that is a superposition of $\tau$ and $\vac$ (vacuum),
so the fusion space in this case is 2-dimensional.
We can represent basis states for this space using diagrams of definite total charge for the Wilson loops, and arbitrary states as linear combinations of these diagrams:
\begin{align*}
\includegraphics[]{pic-fibonacci.pdf}
\end{align*}

For $n$ anyons of type $\tau$, the 
dimension of this space grows asymptotically as $\varphi^n$ for $\varphi=\frac{1+\sqrt{5}}{2}$ 
the golden ratio. Additionally, there is a global degeneracy associated with the topology of the 
manifold on which the anyons reside. We will consider systems with the topology of a torus, 
which for the Fibonacci anyons gives rise to a 2-fold degeneracy.
%For convenience 
%(in particular to minimize finite-size effects in our numerical computations~\cite{Brell2013}), 
This global space is the one in which we will encode, and demonstrate protection of, quantum 
information.

Observables associated to non-intersecting loops commute, and so
a basis for the space can be built from a maximal
set of disjoint, nested loops.
For three anyons, two
possible ways of nesting loops are related %these measurements
via $F$-moves:
\begin{align}
\label{e:fmove}
\includegraphics[]{pic-fmove.pdf}
\end{align}
Here we have shown particles with charges $a$, $b$, and $c$
as well as total charge $d$.
The lefthand side shows a state where the anyons with charge $a$ and $b$ are
observed to have total charge $i$.
The righthand side involves a superposition 
over total charge $j$ of the anyons with charge $b$ and $c$.
The green line serves to linearly
order the anyons, and track
deformations in the underlying space 
(the history of braiding processes).
It also denotes the direction along
which $F$-moves occur.

%Here we show how braids effect observables:
%Here we show the $R$-move, corresponding to a (clockwise) braiding processes:
A clockwise braiding process is represented by an $R$-move:
\begin{align}
\label{e:rmove}
\includegraphics[]{pic-rmove.pdf}
\end{align}
%where $a$, $b$, and $c$ represent particle types, and $F$-moves,
%representing the associativity rule for fusion processes, 
This is meant to show a \emph{half-twist} of the space
inside the observable with value $c$
holding two anyons (punctures) with charges $a$ and $b$.
We are free to draw loops around charges,
and in general braiding will twist these loops in different
ways:
\begin{align*}
\includegraphics[]{pic-braid-loop.pdf}
\end{align*}
%
%
These diagrams can be composed, and the fusion operation replaces the 
interior of a disc by a single charge.
These pictures, abstracted from the quantum context,
are called \emph{curve diagrams} and are used in
the study of braid groups \cite{Dehornoy2002}.
%In general, such a curve diagram is 
%a continuous embedding of a line that passes through 
%each puncture once. 
%Remarkably, any two such linear orderings are related by
%braiding (half-twists).
%\dude{optional paragraph..}

For the Fibonacci anyons, the non-trivial $R$ and $F$ moves are 
\begin{equation*}
	R_{\vac}^{\tau\tau} = \e^{\frac{-4\pi i}{5}} 
	\ \ , \ \
	R_\tau^{\tau\tau}= \e^{\frac{3\pi i}{5}} 
	\ \ , \ \
	F_{\tau}^{\tau\tau\tau} = \begin{pmatrix}\varphi^{-1}&\varphi^{-\frac{1}{2}}\\\varphi^{-\frac{1}{2}}&-\varphi^{-1}\end{pmatrix} \,,
\end{equation*}
where the matrix is given in a basis labelled $(\vac,\tau)$.
%unitary actions on the fusion space. Pair-creation events should be considered isometries 
%that enlarge the fusion space, while fusion events correspond to projective measurements 
%of combined charge of some number of particles.
For detailed description of these 
processes in Fibonacci anyons, see e.g.~Ref.~\cite{Nayak2008} and references therein. 
Braiding of Fibonacci anyons and measurements of fusion outcomes is known to be 
sufficient to implement universal quantum computation~\cite{Freedman2002, Nayak2008}.

This treatment of anyon dynamics we consider to be a phenomenological model;
it neglects any microscopic details of the system.
This is consistent with the principles of topologically 
ordered systems and anyonic physics, where the key universal features describing the 
anyon model correspond to large length-scale physics, while the microscopic physics plays 
a less important (and non-universal) role.

%We use a phenomenological model of Fibonacci anyon dynamics which neglects any 
%microscopic details of the system. 
%This is consistent with the principles of topologically 
%ordered systems and anyonic physics, where the key universal features describing the 
%anyon model correspond to large length-scale physics, while the microscopic physics plays 
%a less important (and non-universal) role.
%Our simulation is in effect that of a $(2+1)$-dimensional topological 
%quantum field theory, which allows for topological processes such as 
%pair-creation, braiding, and fusion of anyonic charges.

@@

% We need to relate this to discrete models: toric code, string nets, etc.

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate connected components 
of average size $O(\log(n))$ and variance $O(1)$~\cite{Bazant2000}.
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected components correspond to sets of anyons that could not have interacted at any 
point in their history. 
We are free to neglect the degrees of freedom associated with braiding between components 
because each component has trivial total charge.
This allows us to simulate the braiding processes within each component separately. 
In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over components. 
Since each 
component has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. 
There are 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in the lattice size $L$~\cite{Grimmett1989}.

However, random noise processes are not the only dynamics that we need to consider. We 
must also consider the effect of the error-correction routine itself. This acts iteratively to fuse 
particles on increasing length scales. While this kind of fusion would typically merge components, 
forcing us to compute dynamics of larger and larger sets of anyons,
large components are sparsely distributed 
(and thus unlikely to be merged), and in addition at each length scale the total number of 
anyons present is dramatically reduced by fusion, leading to a smaller number of anyons that 
must be simulated.

In the regime where the action of noise plus decoder does not percolate it is
clear that the simulation is efficient.
However at some point, with strong enough noise, the state will
no longer decompose at all and computing dynamics will
become exponentially difficult in the system size.
Despite this, we use heuristics %for finite lattice size, we use heuristics
that minimize non-local braid moves, which helps to
maintain the state as a separable product beyond the
sub-percolated regime.
This enables simulation of error correction in
the area around decoder threshold for lattice sizes up to $L=128.$
%However, for finite lattice size, this regime is somewhere
%beyond the decoder threshold, which is the 
%point at which the decoder creates nontrivial (extensive) loops.
%\stf{CGB and I are still not confident in the previous sentence or the next one}
%We can view the system state as a tensor network, on which
%the decoder is performing an (efficiently simulable) contraction.
%This enables simulation of error correction in
%the area around decoder threshold for lattice sizes up to $L=128.$
%\dude{more polish needed...}

@@

There are several interesting avenues for further research. 
Although it is not completely obvious how to do so, 
applying these methods to more realistic models such as concrete 
microscopic spin models or models with non-topological features would give 
direct insight into practical error rates needed for topological quantum memories in non-Abelian systems. 
It would be extremely interesting to find \emph{classical} spin models 
whose phase diagram encodes the threshold for error correction in these systems~\cite{Dennis2002}.
%Another potential direction is to incorporate tensor network methods for 
%simulating anyons~\cite{Pfeifer2012, Singh2014} to improve the speed and scale of the simulations. 
Finally, perhaps the most important open question is whether 
\emph{fault-tolerant} decoders can be designed for nonabelian anyon codes.

>>>>

%------------------------------------------------------------------------------------------------------------%
\paragraph{Topological model.}

@@

We consider a 
2-dimensional system with
point-like excitations or quasiparticles known as anyons.
These anyons show exotic statistics under exchange.
The worldlines of these exchanges form \emph{braids}
in (2+1)-dimensions
which act unitarily on (the Hilbert space of) degrees of freedom known as the \emph{fusion space.}
To access this space we define
the topological observables of the system: these
are total charge measurements within regions bounded by \dude{directed} closed loops (Wilson loops).
%The algebraic properties of these processes
These dynamics and observables obey algebraic rules
given by a unitary modular tensor category~\cite{Wang2010b}. % (UMTC). 
However, from the perspective of
\emph{topological quantum field theory}
anyons are modelled
as punctures in a surface~\cite{Pfeifer2014}, and it is not
the anyons themselves that braid,
but the manifold in which the anyons reside 
that becomes deformed (twisted) by these braid moves.
This is the perspective we adopt (see \Fref{f:braidloop}).
\dude{Fig or FIG ?}
%In particular, this shows the action of braiding
%on arbitrary observables (Wilson loops) as
%these deform with the manifold.

@@

\begin{figure}[t!]
\begin{center}
    \includegraphics[]{pic-braid-loop.pdf}
\caption{
%The perspective of TQFT.
Here we show the world lines of a braid.
%This braid relates two different ways of observing the same charges.
In this case, we see how it relates two different ways of observing (or fusing) the top two charges.
%This braid relates two different ways of observing the same charges.
%We are free to draw these loops around charges,
In general braiding will twist these loops in different ways
according to how the manifold itself is being deformed.
\label{f:braidloop}
}
\end{center}
\vspace{-10pt}
\end{figure}

%The state \dude{space?} of a collection of anyons
%can be described by the observables of the system.
The \dude{key?} difference between abelian and non-abelian
anyon theories is that the outcomes of measurements in an abelian theory
are determined uniquely particle content only.
%The defining characteristic of a \emph{non-abelian}
%anyon theory is that ...
That is, the fusion space is one-dimensional for abelian anyons, and generally larger for non-abelian anyons.
In particular, we consider a system supporting (non-abelian) Fibonacci
anyon excitations, denoted by $\tau.$
Two such anyons can have total charge
that is a superposition of $\tau$ and $\vac$ (vacuum),
so the fusion space in this case is 2-dimensional.
We can represent basis states for this space using diagrams 
of definite total charge for the Wilson loops, 
and arbitrary states as linear combinations of these diagrams:
\begin{align*}
\includegraphics[]{pic-fibonacci.pdf}
\end{align*}
For $n$ anyons of type $\tau$, the 
dimension of this space grows asymptotically as $\varphi^n$ for $\varphi=\frac{1+\sqrt{5}}{2}$ 
the golden ratio.

Observables associated to non-intersecting loops commute, and so
a basis for the space can be built from a maximal
set of disjoint, nested loops.
For three anyons, two
possible ways of nesting loops are related
via $F$-moves:
\begin{align}
\label{e:fmove}
\includegraphics[]{pic-fmove.pdf}
\end{align}
Here we have shown particles with charges $a$, $b$, and $c$
as well as total charge $d$.
The lefthand side shows a state where the anyons with charge $a$ and $b$ are
observed to have total charge $i$.
The righthand side involves a superposition 
over total charge $j$ of the anyons with charge $b$ and $c$.

A clockwise braiding process is represented by an $R$-move:
\begin{align}
\label{e:rmove}
\includegraphics[]{pic-rmove.pdf}
\end{align}
This is meant to show a \emph{half-twist} of the space
inside the observable with value $c$
holding two anyons (punctures) with charges $a$ and $b$.

The green line serves to linearly
order the anyons, and track
deformations in the underlying space 
(the history of braiding processes).
It also denotes the direction along
which $F$-moves occur.

These diagrams can be composed (glued)
by regarding the outermost observable
as a charge in another diagram.
This operation distributes multiplicatively
over superpositions, and it also respects the $R$ and $F$-moves.
Conversely, \emph{fusion} of anyons corresponds to 
replacing the interior of a loop by a single charge.
%and conversely, the fusion operation acts by replacing the 
%interior of a loop by a single charge.

%Abstracted from the quantum context,
%these \emph{curve diagrams} are used in
%the study of braid groups \cite{Dehornoy2002}.
%In general, such a curve diagram is 
%a continuous embedding of a line that passes through 
%each puncture once. 
%Remarkably, any two such linear orderings are related by
%braiding (half-twists).
%\dude{optional paragraph..}

%Pair-creation events should be considered isometries 
%that enlarge the fusion space, while fusion events correspond to projective measurements 
%of combined charge of some number of particles.

For the Fibonacci anyons, the non-trivial $R$ and $F$ moves are 
\begin{equation*}
	R_{\vac}^{\tau\tau} = \e^{\frac{-4\pi i}{5}} 
	\ \ , \ \
	R_\tau^{\tau\tau}= \e^{\frac{3\pi i}{5}} 
	\ \ , \ \
	F_{\tau}^{\tau\tau\tau} = \begin{pmatrix}\varphi^{-1}&\varphi^{-\frac{1}{2}}\\\varphi^{-\frac{1}{2}}&-\varphi^{-1}\end{pmatrix} \,,
\end{equation*}
where the matrix is given in a basis labelled $(\vac,\tau)$.

Finally, there is a global degeneracy associated with the topology of the 
manifold on which the anyons reside. We will consider systems with the topology of a torus, 
which for the Fibonacci anyons gives rise to a 2-fold degeneracy.
This global space is the one in which we will encode, and demonstrate protection of, quantum 
information. 
For detailed description of these 
processes in Fibonacci anyons, see e.g.~Ref.~\cite{Nayak2008} and references therein. 
Braiding of Fibonacci anyons and measurements of fusion outcomes is known to be 
sufficient to implement universal quantum computation~\cite{Freedman2002, Nayak2008}.

This treatment of anyon dynamics we consider to be a phenomenological model;
it neglects any microscopic details of the system.
This is consistent with the principles of topologically 
ordered systems and anyonic physics, where the key universal features describing the 
anyon model correspond to large length-scale physics, while the microscopic physics plays 
a less important (and non-universal) role.
\dude{but how do we relate to discrete models: toric code, string nets, etc.
microscopic lattice models.}

@@

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate connected components 
of average size $O(\log(n))$ and variance $O(1)$~\cite{Bazant2000}.
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected components correspond to sets of anyons that could not have interacted at any 
point in their history. 
We are free to neglect the degrees of freedom associated with braiding between components 
because each component has trivial total charge.
This allows us to simulate the braiding processes within each component separately. 
In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over components. 
Since each 
component has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. 
There are 
random processes that violate this reasoning, but these are suppressed 
exponentially in the lattice size $L$~\cite{Grimmett1989}.

However, random noise processes are not the only dynamics that we need to consider. We 
must also consider the effect of the error-correction routine itself. This acts iteratively to fuse 
anyons on increasing length scales. While this kind of fusion would typically merge components, 
forcing us to compute dynamics of larger and larger sets of anyons,
large components are sparsely distributed
(and thus unlikely to be merged), and in addition at each length scale the total number of 
anyons present is dramatically reduced by fusion, leading to a smaller number of anyons that 
must be simulated.

In the regime where the action of noise plus decoder does not percolate
it is clear \dude{check this} that the simulation is efficient.
However, with strong enough noise the state will
no longer decompose at all and computing dynamics will
become exponentially difficult in the system size.
Despite this, we use heuristics %for finite lattice size, we use heuristics
that minimize non-local braid moves, which helps to
maintain the state as a separable product beyond the
sub-percolated regime.
This enables simulation of error correction in
the area around decoder threshold for lattice sizes up to $L=128.$

@@

There are several interesting avenues for further research. 
Although it is not completely obvious how to do so, 
applying these methods to more realistic models such as concrete 
microscopic spin models or models with non-topological features would give 
direct insight into practical error rates needed for topological quantum memories in non-Abelian systems. 
It would also be interesting to find \emph{classical} spin models 
whose phase diagram encodes the threshold for error correction in these systems~\cite{Dennis2002}.
%Another potential direction is to incorporate tensor network methods for 
%simulating anyons~\cite{Pfeifer2012, Singh2014} to improve the speed and scale of the simulations. 
%Finally, perhaps the most important open question is whether 
Finally, we could ask whether 
\emph{fault-tolerant} decoders can be designed for nonabelian anyon codes.
